A family of accessibility measures derived from spatial interaction principles

### 2025-05-20 paper/manuscript.tex
incorporated rafa's latest comments -- preprint draft ready.

\title{A family of accessibility measures derived from spatial
interaction principles}
\usepackage{etoolbox}
\makeatletter
\providecommand{\subtitle}[1]{% add subtitle to \maketitle
  \apptocmd{\@title}{\par {\large #1 \par}}{}{}
}
\makeatother
\subtitle{Preprint submitted to peer-review in May 2025}
\author{Anastasia Soukhov \and Rafael H. M. Pereira \and Christopher D.
Higgins \and Antonio Páez}
\date{}

\begin{document}
\maketitle
\begin{abstract}
Transportation planning has long prioritized the efficiency of movement,
or mobility. However, the concept of accessibility represents a more
comprehensive evolution, shifting focus from mere movement to the
potential to reach (i.e., spatially interact) with desired destinations.
Despite growing recognition of accessibility-based planning approaches,
the concept remains fragmented, with inconsistent definitions and
unclear interpretations. This work's aim is to clarify and unify the
concept of accessibility by connecting it into spatial interaction
modeling. We demonstrate that widely used mobility and accessibility
models, such as gravity-based accessibility and spatial interaction
models, share common theoretical roots. From this foundation, we define
a family of accessibility measures within the principles of spatial
interaction. In demonstrating these developments, three contributions
emerge: A) the definition and linkages between four members of the
family: specifically, the `unconstrained' measure (i.e., Hansen-type
measure), the `total constrained' measure (i.e., a constrained version
of the Hansen-type measure), the `singly constrained' measure (i.e.,
related to the popular two step floating catchment approach (2SFCA)),
and the `doubly constrained' measure representing realized interactions
or `access', effectively equal to the doubly constrained spatial
interaction model in formulation. B) How to formulaically specify the
total-, singly- and doubly- constrained accessibility measures. C) The
interpretability advantages of the family, as these constrained
accessibility measures yield values in units of the number of potential
``opportunities for spatial interaction'' or ``population for spatial
interaction'' for each zone and zonal flow. By clarifying the concept of
`potential' in accessibility, demonstrating theoretical and formulaic
linkages across popular accessibility and spatial interaction models,
and reintroducing units, this work contributes a family of measures that
unlock a clearer, more interpretable, and cohesive foundation for
accessibility analysis.
\end{abstract}

@@

Accessibility can be seen to represent an evolution of mobility;
accessibility may be defined as the ``potential of opportunities for
{[}spatial{]} interaction'' (Hansen 1959), in contrast to mobility which
is simply, movement in itself or `spatial interaction'. Mobility has
been the basis of traditional transportation planning approaches
(Ortúzar and Willumsen 2011). While accessibility brings a more holistic
understanding of combined transportation and land use systems,
incorporating both the explicit consideration of destination as
opportunity and the concept of \emph{potential} (S. L. Handy and
Niemeier 1997).

The growing interest in accessibility has been accompanied by a boom in
scholarly research using different methods and focusing on different
research contexts; it has grown to include studies of access to
employment (e.g., Karst and Van Eck 2003; Grengs 2010; Páez et al. 2013;
Merlin and Hu 2017; Tao et al. 2020), health care (e.g., Luo and Wang
2003; Páez et al. 2010; Wan, Zou, and Sternberg 2012; Delamater 2013;
Boisjoly, Moreno-Monroy, and El-Geneidy 2017; Pereira et al. 2021),
green spaces (Reyes, Paez, and Morency 2014; Rojas et al. 2016; Liang,
Yan, and Yan 2024), schools (e.g., Williams and Wang 2014; Romanillos
and Garcia-Palomares 2018; Marques, Wolf, and Feitosa 2021), social
contacts (e.g., Neutens et al. 2007; S. Farber, Páez, and Morency 2012;
S. Farber et al. 2013), and regional economic analysis (e.g., R.
Vickerman, Spiekermann, and Wegener 1999; Lopez, Gutierrez, and Gomez
2008; Ribeiro, Antunes, and Páez 2010; Gutierrez et al. 2011) among many
other domains of application. In other words, accessibility analysis is
used today to broadly understand the potential to reach (or spatially
interact with) opportunities that are important to people (Ferreira and
Papa 2020). However, despite its growth in popularity in scholarly
works, challenges remain with respect to the more widespread adoption of
accessibility in planning practice. For instance, the diversity of
accessibility definitions has been flagged by van Wee (2016), S. Handy
(2020), and Kapatsila et al. (2023). Further, difficulties in the
interpretability and communicability of outputs has also been noticed by
Karst T. Geurs and van Wee (2004), van Wee (2016), and Ferreira and Papa
(2020).

Furthermore, the adoption of accessibility in planning practice is not
necessarily made easier when potential adopters have to contend with a
plethora of definitions, each seemingly more sophisticated but less
intuitive than the last (Kapatsila et al. 2023). At a high level, Karst
T. Geurs and van Wee (2004) identify four families of accessibility
measures: infrastructure-, place-, person-, and utility-based. Of the
place-based family, which is the focus of this paper, the menu has grown
to include gravity-based accessibility (e.g., Hansen 1959; Pirie 1979),
cumulative opportunities (e.g., Wachs and Kumagai 1973; Pirie 1979; Ye
et al. 2018), modified gravity (e.g., Schuurman, Berube, and Crooks
2010), 2-Step Floating Catchment Areas (e.g., Luo and Wang 2003),
Enhanced 2-Step Floating Catchment Areas (e.g., Luo and Qi 2009),
3-Stage Floating Catchment Areas (e.g., Wan, Zou, and Sternberg 2012),
Modified 2-Step Floating Catchment Areas (e.g., Delamater 2013), inverse
2-Step Floating Catchment Areas (e.g., F. Wang 2021), and n-steps
Floating Catchment Areas (Liang, Yan, and Yan 2024). How is a
practitioner to choose among this myriad options? What differences in
accessibility scores should matter, and how should they be communicated?
{[}see van Wee (2016); p.~14{]}.

More specifically, this work seeks to address the breath of place-based
accessibility's definitions and lack of interpretability by
demonstrating that mobility-oriented models--such as the commonly used
gravity-based accessibility (e.g., Hansen 1959) and the spatial
interaction model (e.g., Wilson 1971)-- are rooted in the same spatial
interaction modeling foundation. In fact, accessibility can be specified
using spatial interaction principles and a \emph{family of measures},
akin to the family of spatial interaction models introduced in Wilson
(1971), can be defined using balancing factors that constrains values
based on known information about the land-use.

In this way, this work aims to offer three contributions: (1) introduce
and (2) define a family of \emph{constrained} accessibility measures,
namely the total constrained accessibility measure, the singly
constrained accessibility measure, and the doubly constrained measure.
These measures are explicitly connected to popular measures such as the
Hansen-type accessibility (Hansen 1959), the popular competition
approach of the 2-Step Floating Catchment Area (2SFCA) (Q. Shen 1998a;
Luo and Wang 2003), and the concept of market potential, or what Reilly
called `potential trade territories' (Reilly 1929) and further developed
in other examples (C. D. Harris 1954; R. W. Vickerman 1974). These
Wilson-analogous balancing factors introduced are also defined in order
of increasing restrictiveness to shed light on the role of
\emph{potential} in accessibility and access. (3) This work also aims to
demonstrate the intuitive value of restoring units to the resulting raw
accessibility values. Each zonal and zonal flow value from a constrained
accessibility measure is in interpretable units, namely the number of
`opportunities for spatial interaction' or `population for spatial
interaction'. This is in contrast to conventional accessibility measures
that yield values in units of `opportunities weighted by some
representation of travel friction'.

To achieve these stated objectives, this paper contends that
accessibility research must reconnect with its spatial interaction
origins. Particularly, we argue that an important aspect of spatial
interaction modelling--namely, constraining the results to match
empirical observations--was never effectively reincorporated into
accessibility analysis. Empirical constraints were embraced by early
spatial interaction literature following the work of Wilson (1971), but
this stream of literature tended to flow separately from research
inspired by Hansen (1959)' accessibility. The application of Wilson
(1971)`s empirical constraints supported the development of various
spatial interaction models that remain relevant in research and practice
today (Ortúzar and Willumsen 2011). However, the same cannot be said of
the contemporary accessibility literature, where empirical constraints
were not explicitly adopted. We argue that the absence of empirical
constraints (and their attendant proportionality constants) has
contributed to some of accessibility analysis' interpretability issues;
for instance, the fuzziness of insights beyond simple proportional
statements like `higher-than' or `lower-than' (E. J. Miller 2018).
Moreover, without constraints: we lose track of what accessibility
measures, that is, the number of opportunities that can be spatially
interacted with. Without a clear sense of measurement units,
comparability between accessibility measures, across cities, and
transport modes, may also become compromised.

Fortunately, accessibility and spatial interaction modelling literature
share common headwaters, and the latter has given careful attention to
measurement units and their interpretability. It is by looking to the
past that we believe accessibility analysis can newly wade into the
future. We continue this work in the following section by tracing the
development of accessibility from its origins in spatial interaction:
from the Newtonian gravitational expression in Ravenstein (1889) through
to the seminal accessibility work of Hansen (1959). We then present
evidence for a narrative highlighting the marked divergence between
accessibility and spatial interaction modelling research after the work
of Wilson (1971). Next, we hark back to Wilson (1971)'s spatial
interaction models, and use them to derive a family of accessibility
measures based analogous constraints. We illustrate members of this
family of constrained accessibility measures with a simple numerical
example. We then conclude by discussing the uses of these measure and
their interpretation.

@@

Alerted readers will notice that
Equation~\ref{eq-stewart-population-potential-sum}, with some
re-organization of terms, is formally equivalent to our modern
definition of accessibility.

@@

\(S_{i}\) in Equation~\ref{eq-accessibility} is a measure of the
accessibility of site \(i\). This is a function of \(O_j\) (the mass of
opportunities at \(j\)), \(d_{ij}\) (the cost, e.g., distance or travel
time, incurred to reach \(j\) from \(i\)), and \(\beta\) (a parameter
that modulates the friction of cost). Today, Hansen is frequently cited
as the father of modern accessibility analysis (e.g., Reggiani and
Martín 2011), and Hansen-type accessibility is commonly referred to as
the gravity-based accessibility measure.

However, Hansen's use of Stewart's \emph{population potential} measure
included one crucial omission that afflicts the literature to this day.
The omission is that between Stewart (1948) and Hansen (1959), the
proportionality constant \(G\) in
Equation~\ref{eq-stewart-population-potential} vanished. This constant
was not explicitly addressed in Hansen (1959) and accessibility research
continues to evolve without it. Since Hansen (1959), accessibility
analysis has been widely used in numerous disciplines but, to our
knowledge, the proportionality constant has remained implicit or
forgotten, with no notable developments to explicitly acknowledge or
determine it.

The omission of this constant generates a fundamental problem for the
measurement unit of accessibility estimates, which has fundamental
implications for the interpretation, communication and comparability of
accessibility analysis. Those reading Hansen (1959) must recall that
Stewart (1948) had set the proportionality constant \(G\) to 1, with a
note that ``\(G\) {[}was{]} left for future determination: a suitable
choice of other units can reduce it to unity'' {[}p.~34{]}. In practice,
the persistent omission of the constant in accessibility analyses means
that \(G\) continues to be implicitly set to \(1\), even when the
fundamental relationship in accessibility is proportionality (e.g.,
\(S_{i} \propto \sum_j g(O_j)f(d_{ij})\)) and not equality (for
instance, see the formula for accessibility at the top of Figure 1 in Wu
and Levinson 2020). The direct consequence is that without a
proportionality constant, the units of \(S_i\) remain unclear: the unit
of ``potential of opportunity for interaction'' is left free to change
as \(\beta\) is calibrated. For example, if \(c_{ij}\) is distance in
meters, it will be number of opportunities per \(m^{\beta}\) when
\(f(c_{ij}) = d^{-\beta}\) but number of opportunities per
\(e^{-\beta\cdot m}\) when \(f(c_{ij}) = e^{-\beta\cdot m}\). This
undermines the comparability of accessibility metrics with different
decay functions, and renders their results difficult to understand and
communicate. Hansen-style accessibility, therefore, is better thought of
as an ordinal measure of potential that can only be interpreted in terms
of higher and lower accessibility, but which has not palpable meaning
(E. J. Miller 2018).

@@

On the other side of the Atlantic, Alan G. Wilson was developing
related, yet parallel work. In his groundbreaking study (Wilson 1971),
Wilson defined a general spatial interaction model as follows. While
accessibility was characterised as an associated concept of `potential',
the primary focus was on modelling spatial interaction:
\begin{equation}\phantomsection\label{eq-phys-gravity-model}{
T_{ij} = k W_i^{(1)} W_j^{(2)} f(c_{ij})
}\end{equation}

@@

Traditionally, development of the spatial interaction model put an
emphasis on the interpretability of the results (Kirby 1970; Wilson
1967, 1971). But instead of relying on the heuristic of Newtonian
gravity (e.g., some interaction between a mass at \(i\) and a mass at
\(j\) separated by some distance), Wilson's approach was to maximise the
entropy of the system. Entropy maximisation in this case achieves stable
results as a statistical average that represents the population. The
approach works by assuming undifferentiated individual interactions, and
assessing their probabilities of making a particular journey. The result
of Equation~\ref{eq-phys-gravity-model} then is a statistical average
(Wilson 1971; Senior 1979).

@@

In the third case in the family of spatial interaction models is the
production-attraction constrained model. In this case, both
Equation~\ref{eq-constraint1-gravitymodel} and
Equation~\ref{eq-constraint2-gravitymodel} hold simultaneously. These
constraints ensure that the sum of predicted flows from one origin to
all destination, and the predicted flows going to one destination from
all origins must equal the known mass of the origin \(O_i\) and of the
destination \(D_j\). This should hold for all origins and destinations.
The resulting model is, in Wilson's terms, doubly constrained, and takes
the following form:

@@

Derivation of these models is demonstrated in detail elsewhere (e.g.,
Ortúzar and Willumsen 2011; Wilson 1967). It is worth noting, however,
that although Wilson's approach is built on a different conceptual
foundation than the old reference to Newtonian gravity, the work
succeeded at identifying the steps from proportionality to equality to
yield variations of proportionality constants, including the one that
eluded Stewart (1948) and that has been ignored in almost all subsequent
accessibility research. Why was this key element of spatial interaction
models potentially ignored in accessibility research? In the next
section we aim to address this question.

@@

The work of Hansen (1959) and Wilson (1971) responded to important
developments, in particular a need ``to meet the dictates and needs of
public policy for strategic land use and transportation planning''
(Michael Batty 1994). These dictates and needs were far from trivial. In
the United States alone, the Federal-Aid Highway Act of 1956 authorized
the creation of the U.S. Interstate Highway System, with a budget that
ultimately exceeded one hundred billion dollars (Weiner 2016; MDOT
2007). Spatial interaction modelling was incorporated into institutional
modelling practices meant to ``predict and provide'', i.e., predict
travel demand and supply transportation infrastructure (Kovatch, Zames,
et al. 1971; Weiner 2016). Accessibility, at the time, did not quite
have that power, as it did not quantify trips, but rather something
somewhat more elusive: it predicts the less tangible ``potential'' for
spatial interaction with opportunities. In this way, where spatial
interaction modelling became a key element of transportation planning
practice, accessibility remained a somewhat more academic pursuit, and
the two streams of literature only rarely connected.

@@

To illustrate this point, we conducted a bibliographic analysis of the
literature that cites Hansen (1959), Wilson (1971), or both. We
retrieved all relevant documents using the Web of Science ``Cited
References'' functionality, and the digital object identifiers of Hansen
(1959) and Wilson (1971). As a result of this search, we identified
1,788 documents that cite Hansen (1959), only 258 documents that cite
Wilson (1971), and 76 that cite both. The earliest document in this
corpus dates to 1976 and the most recent is from 2025. The number of
documents per year appears in Figure~\ref{fig-docs-per-year}, where we
see the frequency of documents over a span of almost fifty years. In
particular, we notice the remarkable growth in the number of papers that
cite Hansen (1959) compared to those that cite Wilson (1971).

As noted, literature that cites both Wilson (1971) and Hansen (1959) are
sparse (only 3.6\% of the corpus) but highly coupled. However, we can
discern from a reading of these works that they too are divergent, with
one stream focused on developing accessibility (i.e., \emph{potential
for} spatial interaction) and another on spatial interaction. The
focuses of these divergent streams contribute this paper's broader
hypothesis that the concepts of accessibility and spatial interaction
have remained largely disconnected.

On one hand, the stream of literature focused on spatial interaction
models inspired by Wilson (1971) and which cites both Hansen (1959) and
Wilson (1971), tends to contribute to understanding how accessibility is
interpreted and incorporated in spatial interaction models, treating it
as a separate (but related) phenomenon or variable. Specifically, some
early works interpret the spatial interaction model's balancing factors
(Equation~\ref{eq-production-constrained-balancing-factor} or
Equation~\ref{eq-doubly-constrained-balancing-factors}) as the inverse
of Hansen (1959)`s model (B. Harris and Wilson 1978; G. Leonardi 1978;
A. Stewart Fotheringham 1981; A. S. Fotheringham 1985), recognizing it
as a ``common sense'' approach (Morris, Dumble, and Wigan 1979, 99) to
including accessibility in the spatial interaction model, though further
exploration of its relationship is warranted (M. Batty and March 1976).
Some authors have explored this relationship, for instance as in A. S.
Fotheringham (1985) who demonstrates how the spatial interaction model
may insufficiently explain spatial patterns and suggest that explicitly
defining destinations' accessibility as a variable within the model that
may remedy the issue (e.g., the \emph{competition destination} model).
Other works used both Hansen (1959) and Wilson (1971)'s framework in
conjunction, such as in defining location-allocation problems in
operations research (G. Leonardi 1978; Beaumont 1981), estimating trip
flows (or some other spatial interaction flows) alongside accessibility
(e.g., Clarke, Eyre, and Guy 2002; Grengs 2004; Türk 2019), or
considering accessibility within spatial interaction models, in line
with A. S. Fotheringham (1985)'s demonstration (e.g., Beckers et al.
2022). Other works departed from Hansen (1959)'s definition and aligned
with spatial interaction in different ways, such as using micro-economic
consumer behaviour concepts to express potential for spatial interaction
(Morris, Dumble, and Wigan 1979; Giorgio Leonardi and Tadei 1984).

>>>>

\title{A family of accessibility measures derived from spatial
interaction principles}
\usepackage{etoolbox}
\makeatletter
\providecommand{\subtitle}[1]{% add subtitle to \maketitle
  \apptocmd{\@title}{\par {\large #1 \par}}{}{}
}
\makeatother
\subtitle{Submitted preprint for peer-review (May 2025)}
\author{Anastasia Soukhov \and Rafael H. M. Pereira \and Christopher D.
Higgins \and Antonio Páez}
\date{}

\begin{document}
\maketitle
\begin{abstract}
Transportation planning has long prioritized the efficiency of movement,
or mobility. However, the concept of accessibility represents a more
comprehensive evolution, shifting focus from mere movement to the
potential to reach (i.e., spatially interact) with desired destinations.
Despite growing recognition of accessibility-based planning approaches,
the concept remains fragmented, with inconsistent definitions and
unclear interpretations. This work's aim is to clarify and unify the
concept of accessibility by connecting it into spatial interaction
modeling. We demonstrate that widely used mobility and accessibility
models, such as gravity-based accessibility and spatial interaction
models, share common theoretical roots. From this foundation, this paper
offers three contributions: (A) we introduce a family of accessibility
measures within the principles of spatial interaction, and (B) formally
define four members of the family, namely the `unconstrained' measure
(i.e., Hansen-type accessibility), the `total constrained' measure
(i.e., a constrained version of the Hansen-type accessibility), the
`singly constrained' measure (i.e., related to the popular two step
floating catchment approach - 2SFCA), and the `doubly constrained'
measure representing realized interactions or `access', effectively
equal to the doubly constrained spatial interaction model; and (C) we
demonstrate the interpretability advantages of the family, as these
constrained accessibility measures yield values in units of the number
of potential ``opportunities for spatial interaction'' or ``population
for spatial interaction'' for each zone and zonal flow. The family of
accessibility measures proposed here clarifies the concept of
`potential' in accessibility, demonstrates theoretical and formulaic
linkages across popular accessibility and spatial interaction models,
and reintroduces measurement units into accessibility measures. By doing
so, we believe this family of measures can unlock a clearer, more
interpretable, and cohesive foundation for accessibility analysis.
\end{abstract}

@@

Accessibility may be defined as the ``potential of opportunities for
{[}spatial{]} interaction'' (Hansen 1959), in contrast to mobility which
is simply, movement in itself or `spatial interaction'. Mobility has
been the basis of traditional transportation planning approaches
(Ortúzar and Willumsen 2011). While accessibility brings a more holistic
understanding of combined transportation and land use systems,
incorporating both the explicit consideration of destination as
opportunity and the concept of \emph{potential} (S. L. Handy and
Niemeier 1997).

The growing interest in accessibility has been accompanied by a boom in
scholarly research using different methods and focusing on different
research contexts; it has grown to include studies of access to
employment (e.g., Karst and Van Eck 2003; Grengs 2010; Páez et al. 2013;
Merlin and Hu 2017; Tao et al. 2020), health care (e.g., Luo and Wang
2003; Páez et al. 2010; Wan, Zou, and Sternberg 2012; Delamater 2013;
Boisjoly, Moreno-Monroy, and El-Geneidy 2017; Pereira et al. 2021),
green spaces (Reyes, Paez, and Morency 2014; Rojas et al. 2016; Liang,
Yan, and Yan 2024), schools (e.g., Williams and Wang 2014; Romanillos
and Garcia-Palomares 2018; Marques, Wolf, and Feitosa 2021), social
contacts (e.g., Neutens et al. 2007; S. Farber, Páez, and Morency 2012;
S. Farber et al. 2013), and regional economic analysis (e.g., R.
Vickerman, Spiekermann, and Wegener 1999; Lopez, Gutierrez, and Gomez
2008; Ribeiro, Antunes, and Páez 2010; Gutierrez et al. 2011) among many
other domains of application. In other words, accessibility analysis is
used today to broadly understand the potential to reach (or spatially
interact with) opportunities that are important to people (Ferreira and
Papa 2020). However, despite its growth in popularity in scholarly
works, challenges remain with respect to the more widespread adoption of
accessibility in planning practice. For instance, the diversity of
accessibility definitions has been flagged by van Wee (2016), S. Handy
(2020), and Kapatsila et al. (2023). Further, difficulties in the
interpretability and communicability of outputs has also been noticed by
many authores, including Karst T. Geurs and van Wee (2004), van Wee
(2016), and Ferreira and Papa (2020).

The adoption of accessibility in planning practice is not necessarily
made easier when potential adopters have to contend with a plethora of
definitions, each seemingly more sophisticated but less intuitive than
the last (Kapatsila et al. 2023). At a high level, Karst T. Geurs and
van Wee (2004) identify four families of accessibility measures:
infrastructure-, place-, person-, and utility-based. Of the place-based
family, which is the focus of this paper, the menu has grown to include
gravity-based accessibility (e.g., Hansen 1959; Pirie 1979), cumulative
opportunities (e.g., Wachs and Kumagai 1973; Pirie 1979; Ye et al.
2018), modified gravity (e.g., Schuurman, Berube, and Crooks 2010),
2-Step Floating Catchment Areas (e.g., Luo and Wang 2003), Enhanced
2-Step Floating Catchment Areas (e.g., Luo and Qi 2009), 3-Stage
Floating Catchment Areas (e.g., Wan, Zou, and Sternberg 2012), Modified
2-Step Floating Catchment Areas (e.g., Delamater 2013), inverse 2-Step
Floating Catchment Areas (e.g., F. Wang 2021), and n-steps Floating
Catchment Areas (Liang, Yan, and Yan 2024). How is a practitioner to
choose among this myriad options? What differences in accessibility
scores should matter, and how should they be communicated? {[}see van
Wee (2016); p.~14{]}.

Here, we seek to address the breath of place-based accessibility's
definitions and lack of interpretability by demonstrating that
mobility-oriented models--such as the commonly used gravity-based
accessibility (e.g., Hansen 1959) and the spatial interaction model
(e.g., Wilson 1971)-- are rooted in the same spatial interaction
modeling foundation. In fact, we propose that accessibility can be
specified using spatial interaction principles as a \emph{family of
measures}, akin to the family of spatial interaction models introduced
in Wilson (1971) that can be defined using balancing factors that
constrains values based on known information about the land-use.

This work aims to offer three contributions: (1) it introduces a family
of accessibility measures within the principles of spatial interaction;
(2) it formally defines three \emph{constrained} accessibility measures
by reintroducing Wilson-analogous balancing factors. These measures are
the total constrained accessibility measure, the singly constrained
accessibility measure, and the doubly constrained measure, which are
explicitly connected to popular measures such as the Hansen-type
accessibility (Hansen 1959), the popular competition approach of the
2-Step Floating Catchment Area (2SFCA) (Q. Shen 1998a; Luo and Wang
2003), and the concept of market potential (C. D. Harris 1954; R. W.
Vickerman 1974). These Wilson-analogous balancing factors introduced are
also defined in order of increasing restrictiveness to shed light on the
role of \emph{potential} in accessibility and access. (3) This work also
demonstrates that the introduction of balancing factors makes
accessibility measures easier to interpret and to communicate by
restoring the measurement units to the resulting raw accessibility
values. Each zonal and zonal flow value from a constrained accessibility
measure is always in interpretable units, namely the number of
`opportunities for spatial interaction' or `population for spatial
interaction'. This is in contrast to conventional accessibility
measures, particularly gravity based measures, that yield values in
units of `opportunities weighted by some representation of travel
friction'.

To achieve these stated objectives, this paper contends that
accessibility research must reconnect with its spatial interaction
origins. Particularly, we argue that an important aspect of spatial
interaction modelling--namely, constraining the results to match
empirical observations--was never effectively reincorporated into
accessibility analysis. Empirical constraints were embraced by early
spatial interaction literature following the work of Wilson (1971), but
this stream of literature tended to flow separately from research
inspired by Hansen (1959)' accessibility. The application of Wilson
(1971)`s empirical constraints supported the development of various
spatial interaction models that remain relevant in research and practice
today (Ortúzar and Willumsen 2011). However, the same cannot be said of
the contemporary accessibility literature, where empirical constraints
were not explicitly adopted. We argue that the absence of empirical
constraints (and their attendant proportionality constants) has
contributed to some of accessibility analysis' interpretability issues;
for instance, the fuzziness of insights beyond simple proportional
statements like `higher-than' or `lower-than' (E. J. Miller 2018).
Moreover, without constraints we lose track of what accessibility
measures, that is, the number of opportunities that can be spatially
interacted with. Without a clear sense of measurement units,
comparability between accessibility measures, across cities, and
transport modes, may also become compromised.

Fortunately, accessibility and spatial interaction modelling literature
share common headwaters, and the latter has given careful attention to
measurement units and their interpretability. It is by looking to the
past that we believe accessibility analysis can newly wade into the
future. We continue this work in the following section by tracing the
development of accessibility from its origins in spatial interaction:
from the Newtonian gravitational expression in Ravenstein (1889) through
to the seminal accessibility work of Hansen (1959). We then present
evidence for a narrative highlighting the marked divergence between
accessibility and spatial interaction modelling research after the work
of Wilson (1971). Next, we hark back to Wilson (1971)'s spatial
interaction models, and use them to derive a family of accessibility
measures based on analogous constraints. We illustrate members of this
family of constrained accessibility measures with a simple numerical
example. We then conclude by discussing the uses of these measure and
their interpretation.

@@

Alerted readers will notice that
Equation~\ref{eq-stewart-population-potential-sum}, with some
re-organization of terms, is formally equivalent to our modern
definition of accessibility popularized by Hansen (1959) in the late
1950s.

@@

\(S_{i}\) in Equation~\ref{eq-accessibility} is a measure of the
accessibility from site \(i\). This is a function of \(O_j\) (the mass
of opportunities at \(j\)), \(d_{ij}\) (the cost, e.g., distance or
travel time, incurred to reach \(j\) from \(i\)), and \(\beta\) (a
parameter that modulates the friction of cost). Today, Hansen is
frequently cited as the father of modern accessibility analysis (e.g.,
Reggiani and Martín 2011), and Hansen-type accessibility is commonly
referred to as the gravity-based accessibility measure.

However, Hansen's use of Stewart's \emph{population potential} measure
included one crucial omission that afflicts the literature to this day.
The omission is that between Stewart (1948) and Hansen (1959), the
proportionality constant \(G\) in
Equation~\ref{eq-stewart-population-potential} vanished. This constant
was not explicitly addressed in Hansen (1959) and accessibility research
continues to evolve without it. Since Hansen (1959), accessibility
analysis has been widely used in numerous disciplines but, to our
knowledge, the proportionality constant has remained forgotten, with no
notable developments to explicitly acknowledge or determine it.

The omission of this constant generates a fundamental problem for the
measurement unit of accessibility estimates, which undermines the
interpretation, communication and comparability of accessibility
analysis. Those reading Hansen (1959) must recall that Stewart (1948)
had set the proportionality constant \(G\) to 1, with a note that
``\(G\) {[}was{]} left for future determination: a suitable choice of
other units can reduce it to unity'' {[}p.~34{]}. In practice, the
persistent omission of the constant in accessibility analyses means that
\(G\) continues to be implicitly set to \(1\), even when the fundamental
relationship in accessibility is proportionality (e.g.,
\(S_{i} \propto \sum_j g(O_j)f(d_{ij})\)) and not equality (for
instance, see the formula for accessibility at the top of Figure 1 in Wu
and Levinson 2020). The direct consequence is that without a
proportionality constant, the units of \(S_i\) remain unclear: the unit
of ``potential of opportunity for interaction'' is left free to change
as \(\beta\) is calibrated. For example, if \(c_{ij}\) is distance in
meters, it will be number of opportunities per \(m^{\beta}\) when
\(f(c_{ij}) = d^{-\beta}\) but number of opportunities per
\(e^{-\beta\cdot m}\) when \(f(c_{ij}) = e^{-\beta\cdot m}\). This
undermines the comparability of accessibility metrics with different
decay functions, and renders their results difficult to understand and
communicate. The Hansen-style accessibility estimates found in the
literature, therefore, are better thought of as ordinal measures of
potential that can only be interpreted in terms of higher and lower
accessibility, but which has not palpable meaning (E. J. Miller 2018).

@@

On the other side of the Atlantic, Alan G. Wilson was developing
related, yet parallel work. In his groundbreaking study (Wilson 1971),
Wilson defined a general spatial interaction model as follows. While
accessibility was characterised as an associated concept of `potential',
the primary focus was on modelling observed spatial interaction:
\begin{equation}\phantomsection\label{eq-phys-gravity-model}{
T_{ij} = k W_i^{(1)} W_j^{(2)} f(c_{ij})
}\end{equation}

@@

Traditionally, the development of the spatial interaction model put an
emphasis on the interpretability of the results (Kirby 1970; Wilson
1967, 1971). But instead of relying on the heuristic of Newtonian
gravity (e.g., some interaction between a mass at \(i\) and a mass at
\(j\) separated by some distance), Wilson's approach was to maximise the
entropy of the system. Entropy maximisation in this case achieves stable
results as a statistical average that represents the population. The
approach works by assuming undifferentiated individual interactions, and
assessing their probabilities of making a particular journey. The result
of Equation~\ref{eq-phys-gravity-model} then is a statistical average
(Wilson 1971; Senior 1979).

@@

The third case in the family of spatial interaction models is the
production-attraction constrained model. In this case, both
Equation~\ref{eq-constraint1-gravitymodel} and
Equation~\ref{eq-constraint2-gravitymodel} hold simultaneously. These
constraints ensure that the sum of predicted flows from one origin to
all destination, and the predicted flows going to one destination from
all origins must equal the known mass of the origin \(O_i\) and of the
destination \(D_j\). This should hold for all origins and destinations.
The resulting model is, in Wilson's terms, doubly constrained, and takes
the following form:

@@

Derivation of these models is demonstrated in detail elsewhere (e.g.,
Ortúzar and Willumsen 2011; Wilson 1967). It is worth noting, however,
that although Wilson's approach is built on a different conceptual
foundation than the old reference to Newtonian gravity, the work
succeeded at identifying the steps from proportionality to equality to
yield variations of proportionality constants, including the one that
eluded Stewart (1948) and that has been overlooked in almost all
subsequent accessibility research. Why was this key element of spatial
interaction models potentially ignored in accessibility research? In the
next section we aim to address this question.

@@

The work of Hansen (1959) and Wilson (1971) responded to important
developments, in particular a need ``to meet the dictates and needs of
public policy for strategic land use and transportation planning''
(Michael Batty 1994). These dictates and needs were far from trivial. In
the United States alone, the Federal-Aid Highway Act of 1956 authorized
the creation of the U.S. Interstate Highway System, with a budget that
ultimately exceeded one hundred billion dollars (equivalent to over
\$600 billion in 2023) (Weiner 2016; MDOT 2007). Spatial interaction
modelling was incorporated into institutional modelling practices meant
to ``predict and provide'', i.e., predict travel demand and supply
transportation infrastructure (Kovatch, Zames, et al. 1971; Weiner
2016). Accessibility, at the time, did not quite have that power, as it
did not quantify trips, but rather something somewhat more elusive: it
predicts the less tangible ``potential'' for spatial interaction with
opportunities. In this way, where spatial interaction modelling became a
key element of transportation planning practice, accessibility remained
a somewhat more academic pursuit, and the two streams of literature only
rarely connected.

@@

To illustrate this point, we conducted a bibliographic analysis of the
literature that cites Hansen (1959), Wilson (1971), or both. We
retrieved all relevant documents using the Web of Science ``Cited
References'' functionality, and the digital object identifiers of Hansen
(1959) and Wilson (1971). As a result of this search, we identified
1,788 documents that cite Hansen (1959), only 258 documents that cite
Wilson (1971), and 76 that cite both. The earliest document in this
corpus dates to 1976 and the most recent is from 2025. The number of
documents per year appears in Figure~\ref{fig-docs-per-year}, where we
see the frequency of documents over a span of almost fifty years. In
particular, we notice the remarkable growth in the number of papers that
cite Hansen (1959) compared to those that cite Wilson (1971) since the
year 2000.

As noted, literature that cites both Wilson (1971) and Hansen (1959) are
sparse (only 3.6\% of the corpus, visualised in pink in
Figure~\ref{fig-docs-per-year}). After reading the works, we can also
discern that they too are divergent, with one stream focused on
developing accessibility (i.e., \emph{potential for} spatial
interaction) and another on spatial interaction. The focuses of these
divergent streams contribute this paper's broader hypothesis that the
concepts of accessibility and spatial interaction have remained largely
disconnected, and at times, improperly conflated.
